---
title: "CHL5233: Quarto in-class assignment"
author: "Daniel Guo"
subtitle: "Countries by region and sub-region"
fontsize: 10pt
format:
  pdf:
    pdf-engine: xelatex
    geometry:
      - top=30mm
      - left=20mm
      - heightrounded
    colorlinks: true
execute:
  warning: false
  message: false
---

```{r}
#| label: setup
#| include: false
library(tidyverse)
library(knitr)
library(kableExtra)
```

## Read in data and create an analytical data set

The supplied files describe the study country list and geographic classifications.
The files are stored in `data/raw/` in this project; paths below are relative to
this report in `reports/`.

```{r}
#| label: read-countries
#| echo: true
countries <- read.table("data/raw/countries.txt", header = TRUE)
dim(countries)
```

The countries data set contains **`r nrow(countries)` countries** and
**`r ncol(countries)` columns**. The supplied country list contains 186 distinct ISO
codes. This report uses all countries in the supplied file.

```{r}
#| label: read-regions
regions <- read.table("data/raw/regions.txt", header = TRUE)
dim(regions)
```

The regions data set contains **`r nrow(regions)` country or territory entries**
and **`r ncol(regions)` columns**.

We match the three-letter ISO codes, avoiding differences in country names.
`semi_join()` retains only entries in `regions` whose codes occur in `countries`.

```{r}
#| label: analytical-data
study_regions <- regions |>
  semi_join(countries, by = c("country_code" = "ISO"))

nrow(study_regions)
```

```{r}
#| include: false
# Check unique identifiers, complete matching and region labels.
stopifnot(
  !anyDuplicated(countries$ISO),
  !anyDuplicated(regions$country_code),
  all(countries$ISO %in% regions$country_code),
  nrow(study_regions) == nrow(countries),
  all(!is.na(study_regions$region) & study_regions$region != ""),
  all(!is.na(study_regions$sub_region) & study_regions$sub_region != "")
)
nrow(study_regions)
```

All **`r nrow(study_regions)` countries** matched a geographic entry; none were
lost or duplicated in the match.

\newpage

## Tabulate countries by region and sub-region

```{r}
#| label: region-table
region_counts <- study_regions |>
  group_by(region) |>
  summarise(count = n(), .groups = "drop") |>
  arrange(desc(count), region)

kable(region_counts, col.names = c("Region", "Count"),
      caption = "Study countries by region", booktabs = TRUE, label = NA) |>
  kable_styling(latex_options = "HOLD_position")
```

\newpage

### Countries by sub-region

```{r}
#| label: sub-region-table
subregion_counts <- study_regions |>
  group_by(region, sub_region) |>
  summarise(count = n(), .groups = "drop") |>
  arrange(desc(count), sub_region)

kable(select(subregion_counts, sub_region, count),
      col.names = c("Sub-region", "Count"),
      caption = "Study countries by sub-region", booktabs = TRUE, label = NA) |>
  kable_styling(latex_options = "HOLD_position")
```

```{r}
#| include: false
stopifnot(sum(region_counts$count) == nrow(countries),
          sum(subregion_counts$count) == nrow(countries))
```

Both tables total **`r sum(region_counts$count)` countries**. Africa has the
largest regional count (54), and Sub-Saharan Africa has the largest
sub-regional count (48).

\newpage

## Data visualization using ggplot

The regional chart summarizes the overall distribution. Each bar counts
individual countries in the analytical data set.

```{r}
#| label: region-chart
#| fig-width: 7
#| fig-height: 3.5
#| fig-cap: "Number of study countries in each region."
ggplot(study_regions,
       aes(x = fct_infreq(region))) +
  geom_bar(fill = "#287D8E", width = 0.7) +
  labs(x = "Region", y = "Number of countries") +
  scale_y_continuous(breaks = seq(0, 60, 10)) +
  theme_minimal(base_size = 11) +
  theme(panel.grid.major.x = element_blank())
```

\newpage

The sub-regional chart uses horizontal bars, colored by region. Reversing the
frequency ordering places the largest count at the top.

```{r}
#| label: sub-region-chart
#| fig-width: 7.2
#| fig-height: 4.8
#| fig-cap: "Study countries by sub-region, ordered by count and colored by region."
ggplot(study_regions,
       aes(x = fct_rev(fct_infreq(sub_region)), fill = region)) +
  geom_bar(width = 0.75) +
  coord_flip() +
  scale_fill_manual(values = c(
    Africa = "#0072B2", Americas = "#D55E00", Asia = "#009E73",
    Europe = "#CC79A7", Oceania = "#E69F00"
  )) +
  scale_y_continuous(breaks = seq(0, 50, 10)) +
  labs(x = "Sub-region", y = "Number of countries", fill = "Region") +
  theme_minimal(base_size = 10) +
  theme(legend.position = "bottom",
        panel.grid.major.y = element_blank())
```

These counts describe geographic representation in the supplied list, rather
than the number or severity of armed conflicts.
